MASCARA: coexpression analysis in data from designed experiments

### 2023-12-11 R/Paper_markdown.Rmd
231211

<!-- Need to change this as it is repeated in introduction -->

@@

# Introduction
<!-- DE or CoE to understand system + feature selection -->
Plant transcriptome studies typically involve the use of designed experiments which aim to induce variation in pathways of interest (POIs) by controlling one or more experimental factors. This can be done through, for example, varying the level of certain essential nutrients, or knocking out/down a gene within a POI. To explore which pathways are affected by the intervention, gene expression can be measured through RNA sequencing (RNAseq). Analysis of RNAseq data generally adheres to one of two approaches: differential expression analysis (DE; @andersDifferentialExpressionAnalysis2010; @robinsonEdgeRBioconductorPackage2010) or coexpression analysis (CoE; @langfelderWGCNAPackageWeighted2008; @tzfadiaCoExpNetVizComparativeCoExpression2016). DE is used to determine which genes are expressed differently between two or more experimental conditions, highlighting known or novel pathways affected by the experimental conditions. CoE, on the other hand, aims at discovering genes that are part of a partially characterized pathway. Here, the expression profiles of known genes of the POI (termed "baits") are used to detect novel pathway members with similar expression patterns (here termed "spikes").

```{r nettype, fig.cap= c("Pathway types exemplified, POI; pathway of interest, RP; related pathway, URP; unrelated pathway. A represents hypothetical biological pathways, B highlights the problem with detecting these pathways when dealing with data from designed experiments.")}
#load network types image

a <- knitr::include_graphics("networktypes.png")
a

Figure \@ref(fig:nettype) outlines pathway nomenclature that will be used throughout this text. *A* represents a true hypothetical pathway including a pathway of interest (POI), related pathway (RP) and an unrelated pathway (URP). All three types of pathway are assumed to be affected to varying degrees by the experimental design. When dealing with experimentally designed data, we have to consider two key sources of variance: between-group variance and within-group variance. Between-group variance pertains to the differences between experimental groups (or conditions). This type of variance is of interest when conducting DE studies. Within-group variance, on the other hand, is the variance within experimental groups, this is not interesting for DE studies but may contain information useful to CoE. It is the excessive between-group variance that leads to the densely connected nature of *B* from \@ref(fig:nettype). Figure \@ref(fig:withinbetween) exemplifies between-group and within-group driven correlations. The plots show the relationship between two genes of the POI (*A* and *C*) and the relationship between two genes that are not part of the same pathway (*B* and *D*).  The total correlations of both pairs of transcripts are high but the second pair’s correlation is driven by the between-group variation and not by correlations within experimental groups. The within-group correlation is lower than the total in both cases, but there is a clear trend in one condition (facet *C*: P- condition in red) for the first gene pair. Because of this, the within group correlation (r) is higher for the first transcript pair (*C*) than the second (*D*), which makes it possible to discern between genes within a POI or RP and genes that both respond similarly to a treatment (URP). Hence, we need to distinguish *total CoE* which is calculated on the total variance and *within-group CoE* where the focus is on the shared variance within experimental groups.

<!-- From a plant biology point of view, the between-group variance and the within-group variance can be quite different and related to different parts of the biological system. It is, in fact, expected that there are mechanistic changes that arise from the very different conditions that could alter the coexpression of genes.  -->

<!-- , examples include: @miettinenSecoiridoidPathwayCatharanthus2014;.  -->

@@

<!-- This phenomenon is well documented in @caldanaHighdensityKineticAnalysis2011 where an experiment was performed investigating the effects of temperature and light on arabidopsis plants over time. In this study the authors investigated variable relationships between metabolites and showed these were strongly dependent on the specific levels of the experimental factors, e.g., shikimic acid and phenylalanine were positively correlated at 21 degrees in high light condition, but negatively correlated at 32 degrees in the dark.  -->

<!-- Between experimental groups different pathways are active, whereas within groups more general (undocumented/uncontrolled) ambient factors such as temperature, light intensity and soil water content can affect the plant and its transcriptome in a concerted fashion. In experiments where the intervention causes large between-group differences to many genes, not making the distinction between types of variance can lead to results that contain total CoE associations driven by similar differences between treatment groups.  -->

@@

<!-- ## Dataset: Experimental Design -->
```{r, echo = FALSE, include=FALSE}
META <- as.data.frame(meta[,1:2])
META$growth_condition[which(META$growth_condition == "1")] <- "P+"
META$growth_condition[which(META$growth_condition == "-1")] <- "P-"

@@

What is notable here is the replicate specific patterns that are apparent amongst the SL genes and largely absent between the SLs (purple) and the non-SL differential genes (green). For example, at time point 3,  most SL genes are expressed lowest in replicate 1 and highest in replicate 3, while at time point 7 replicate 2 has a higher SL gene expression. These within-group patterns are either not shared at all between the SL genes and the other DE genes or to a much lesser extent. This is due to the other genes not directly being part of the same pathway. There are of course other pathways and thus there are still some subtle replicate patterns in the non-SL genes but this effect is much clearer within our pathway of interest. Figure \@ref(fig:ASCA-1) shows the output of the ASCA decomposition in which the PCA scores and loadings of the combined effect of the  Growth condition factor and the interaction of Growth condition and Time is explored. The score plot shows that the replicates of each condition are clustered while the difference between the conditions is much larger. The loading plot shows that although the 9 SL genes are mostly on the outside of the plot, many other genes are as well. Thus many differentially expressed genes are found important that are not part of the SL pathway, in other words ASCA does not detect spike SL genes from others that are similarly or more differentially expressed but not part of the SL pathway.

This marks the goal of this paper, which is to detect within-group coexpressed genes with a set of baits preferentially to other differentially expressed genes that are not part of the same pathway (with no within-group structure).

@@

```{r Dataset1-Overview, echo = F, fig.cap=c("Real data overview. RNAseq of rice root. Heatmap of SL pathway and example highly differential genes.")}

@@

```{r ASCA-1, echo = F, out.width = '100%', fig.cap=c("Scores and loadings of the ASCA model of the combined effect (condition + condition:time interaction).")}
knitr::include_graphics("rice_s_l.png")

@@

The simulations are built upon the results of the ASCA analysis of the combined effect of growth condition and the interaction of growth condition and time. The simulated data is created by simulating score and loading values (as in Figure \@ref(fig:simOverview)) for the combined effect ($\alpha + \alpha\beta$) ) explaining the difference in time profiles of the P+ and P- condition and the main time effect $\beta$, as well as, structured ambient variance ($S$). This is to mimic variance due to gene expression as a function of growth condition and time as well as variance due to coexpression of genes in the same  pathway on top of this experimentally induced DE variance.

@@

The first simulation aims to determine how well the methods perform to find coexpressed genes in the presence of a dominant between-group variance structure and increasing numbers of non-pathway differential genes (genes with high between-group correlation but no within-group correlation). The expression levels of differential transcripts is modeled with the scores and loadings of the growth condition and its interaction with time ( $\mathbf{T}_{\alpha + \alpha\beta}$ and $\mathbf{P}_{\alpha + \alpha\beta}^T$ ). The disturbing time effect is modeled with scores and loadings ( $\mathbf{T}_{\beta}$  and $\mathbf{P}_{\beta}^T$ ). Also included is some random measurement noise, this is modeled as a matrix $\mathbf{E}$ with normally distributed random values $e_{ij} = \mathcal{N}(0,\sigma^2)$.

On top of the between-group effect there is an additional a correlation structure applied to the baits and spikes to create a model of coexpression that is independent to the variance induced by the experimental factors ($\mathbf{X}_S = \mathbf{T}_S\mathbf{P}_S^T$). In this simulation we have a set of 4 baits that have positive loadings for the combined effect along with a variable number of other genes - "Combined Effect Responders". For the structured ambient variance; 2 random sample offset vectors in \(\bm{T}_S\) (each column representing an effect caused by ambient conditions) which our baits and spikes contribute to (i.e. green and black bars in Figure \@ref(fig:simOverview) are non zero in \(\bm{P}_S\)). Similarly to the real data, the gene expression values for each experimental condition are simulated with three independent replicates per time-condition-combination. Each simulated data $\mathbf{Y}$ set thus consists of 24 samples and 2000 genes.

@@

The scores matrices $\mathbf{T}_{\alpha + \alpha\beta}$ and $\mathbf{T}_{\beta}$ each consist of two orthogonal score vectors and their relative levels are indicated by the colour. The different colours in Figure \@ref(fig:simOverview) represent that the scores are orthogonal to each other. Each small block consists of the same score value for each of the replicates within that condition. Loadings $\mathbf{P}_{\alpha + \alpha\beta}^T$ and $\mathbf{P}_{\beta}^T$  are simulated to create the baits and spikes. As displayed in Figure \@ref(fig:simOverview), the green bar represents 4 baits (or known pathway genes), the black bar represents the 12 spikes (“unknown” genes of interest), yellow represents differentially expressed genes not in SL pathway, and purple; a set of differential genes not in SL pathway with an alternative treatment response.

In $\mathbf{P}_{\beta1}^T$ the baits, spikes and other DE genes along with the remaining first half have similar positive values (represented by light grey) where the other 1000 have positive values in $\mathbf{P}_{\beta2}^T$ (grey), these are created to emulate the reality of the real data in which many genes are slightly correlated with eachother through the course of development, as well as to account for some nonlinearities which are otherwise ignored by our model. The remaining elements of $\mathbf{P}_{\alpha + \alpha\beta}^T$ and $\mathbf{P}_{\beta}^T$ are populated with zeroes. The $\mathbf{T}$ matrices represent the simulated scores for their respective factors. For blues, reds, greys and golds we have 4 different shades representing 4 unique values per factor level (total 8 unique values where reds and golds are sign flipped blues and greys respectively). Similarly oranges and pinks represent a growth condition independent time profile with four unique values. $\mathbf{E}$ represents random unstructured technical noise and the additional $\bm{1\mu}^T$ term in \@ref(eq:Sim2-bp) represents a mean offset for each gene.

```{r simOverview, echo=FALSE, fig.cap="Coexpression simulation, simplified overview. Colours correspond to the same genes in the different loading vectors (*P*). Black; baits (known SL genes), green; spikes (unknown SL genes), yellow; other non-pathway DE genes, purple non-pathway DE genes with a different profile. In the score vectors (T); Blues, reds, greys and golds represent experimental conditions, oranges and pinks; time. Each of the textures in the ambient/residual part represent a distinct normally distributed set of random values.", out.width = '100%'}
knitr::include_graphics("Sim_SL_diagram_simplified.png")
```

\begin{equation}
\begin{aligned}
\bm{Y} &= \bm{1\mu}^T + \bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T + \bm{T}_{\beta}\bm{P}_{\beta}^T + \bm{T}_{S}\bm{P}_{S}^T + \bm{E} \\
\text{where for each element ik or jk:}\\
&\bm{T}_{S_{ik}} \sim \mathcal{N}(0,1), \quad \bm{P}_{S_{jk}}^{\text{baits,spikes}} \sim \mathcal{N}(0.5,0.2) \\
\text{and:} &\\
\bm{P}_{S}^{\text{rest}} &= 0
\end{aligned}(\#eq:Sim2-bp)
\end{equation}

<!-- ### Combined effect size and number of differential genes -->

<!-- \begin{equation} -->
<!-- \begin{aligned} -->
<!-- \bm{Y} &= \bm{1\mu}^T + d(\bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T) + \bm{T}_{\beta}\bm{P}_{\beta}^T + \bm{T}_{S}\bm{P}_{S}^T + \bm{E} \\ -->
<!-- \text{where:} &\\ -->
<!-- \bm{P}_{\alpha + \alpha\beta}^T = \begin{bmatrix} -->
<!--        \bm{p}_{11} & \bm{p}_{12} \quad \bm{p}_{13} \\ -->
<!--        \bm{p}_{21} & \bm{p}_{22} \quad \bm{p}_{23} -->
<!--      \end{bmatrix} -->
<!-- \end{aligned}(\#eq:nCER) -->
<!-- \end{equation} -->

The issue to address is the presence of many non-pathway differential genes; those that have strong between group correlations but no within group correlation. Here the differential genes relate  to the yellow bar in \@ref(fig:simOverview). We incrementally increase the number of non-zero values and decrease the number of 0 values, which in figure \@ref(fig:simOverview) relates to increasing the size of the yellow bar, as well as varying the F-norm of the combined effect matrix through the scalar parameter *d* below;

@@

<!-- \(\bm{p}_{2}\) remains the same with \(\bm{p}_{21}\) and \(\bm{p}_{23}\) being 0 and \(\bm{p}_{22}\) (purple bar in \@ref(fig:simOverview)) positive. -->

@@

In order to evaluate the generalisation of methods and indicate the levels of variance required for optimal performance we conduct a larger scale assay. Here we fix the number of non-pathway DE genes to 30 which corresponds (to the yellow bar size) and the size of the within correlation effect matrix (\(\|\bm{T}_S\bm{P}_S^T\|_F = 0.025\)). We investigate a range of values for sizes of between variance (*d*) and random noise (*l*) across random structures (which correspond to repeated simulations of randomised values for $\mathbf{T}$ matrices). Parameters were adjusted according to \@ref(eq:Test-2) on 20 randomised $\mathbf{T}$ matrices.

@@

```{r, echo = FALSE}
# X_funced_classic <- Create_Core(nreps = 3, meta = meta, irr_spikes = TRUE,
#                         struc_resid = FALSE, noise_sd = 0.75, plot = TRUE)   ## sum(X_funced[[1]] != X_core) = 0 with these parameters

@@

# alphaN <- 2
# betaN <- 4
# nreps <- 5
# 
# meta2 <- cbind.data.frame(rep(c(1,-1), each = betaN*nreps), rep(c(1:betaN), each = nreps), rep(c(-1:-betaN,1:betaN), each = nreps))
# 
# colnames(meta2) <- c("growth_condition","time","interaction")

@@

Pearson's correlation coefficient is calculated between all columns of \(\bm{Y}\). 
For example the correlation between a SL gene $j_{sl}$ and a non-SL gene $j_q$ is calculated as:

\begin{equation}
\begin{aligned}
  r_{j_{sl}j_{q}} = \frac{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl}) (y_{ij_{q}} - \bar{y}_j_{q})}{\sqrt{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl})^2 \sum_{i=1}^{I} (y_{ij_{q}} - \bar{y}_j_{q})^2}}
\end{aligned}(\#eq:Pearson)
\end{equation}

For ranked correlations *r* is calculated between each bait and all other genes, with the lowest rank for the highest correlation. Then for each bait the other genes are ranked based on *r*. These ranks are averaged across all baits, similarly to @poretskyMutRankShinyWebapplication2020.

@@

<!-- This approach provides a general solution to build the ASCA effect matrices for any fixed effects ANOVA model. -->

### PLS regression - canonical mode

Partial Least Squares canonical mode (PLS2) is a multivariate regression technique used for modeling relationships between two sets of variables, \( \mathbf{Y} \) and \( \mathbf{B} \).

After splitting the data into the response matrix (containing only baits) \( \mathbf{B} \) (the green bar in the loadings) and the predictor matrix (containing the rest of the data) \( \mathbf{Y} \), resulting in the model of the data:

\begin{equation}
\begin{aligned}
\mathbf{Y} = \mathbf{T}\mathbf{P}^T + \mathbf{E}_Y \\
\text{and} \\
\mathbf{B} = \mathbf{T}\mathbf{Q}^T + \mathbf{E}_B \\
\end{aligned}
\end{equation}

This model is estimated in our work as follows:

<!-- 2. Center and scale both \( \mathbf{X} \) and \( \mathbf{Y} \). -->
1. Initialize \( \mathbf{Y}_0 \) to \( \mathbf{Y} \) and \( \mathbf{B}_0 \) to \( \mathbf{B} \).
2. Iteratively perform the following steps for \( k = 1, 2, \ldots \):
   a. Compute the singular value decomposition (SVD) of \( \mathbf{Y}_{k}^{T} \mathbf{B}_{k} \) to obtain the left and right singular vectors \( \mathbf{u}_k \) and \( \mathbf{v}_k \) respectively.
   b. Project \( \mathbf{Y}_{k} \) and \( \mathbf{B}_{k} \) onto the first left and right singular vectors (\( \mathbf{u}_k \), \( \mathbf{v}_k \)) to obtain the scores \( \bm{t}_k = \mathbf{Y}_{k} \mathbf{u}_k \) and \( \bm{w}_k = \mathbf{B}_{k} \mathbf{v}_k \).
   c. Perform linear regression of \( \mathbf{Y}_k \) and \( \mathbf{B}_k \) onto their corresponding scores \( \bm{t}_k \) and \( \bm{w}_k \) to obtain the respective loadings \( \bm{p}_k \), and \( \bm{q}_k \).
   d. Deflate \( \mathbf{Y}_k \) and \( \mathbf{B}_k \) by removing the variance explained by the $k^{th}$ pair of scores (\( \bm{t}_k \), \( \bm{w}_k \)), and use the deflated matrices as inputs for the next iteration.

PLS2 aims to maximize the covariance between the scores \( \bm{t}_k \) and \( \bm{w}_k \) at each iteration. In the following simulations and application the number of components calculated by PLS2 is set to 2.

<!-- although there are methods to optimally select this number, such as; (REFS). -->

@@

$\text{VIP}_{j} = \sqrt{\frac{ \sum_{k=1}^{K} \mathbf{h}^2_{jk} \cdot \text{VE}_{k}}{\text{VE}_{total} \cdot K}}$

Where k=1...K represents the number of components in the respective models, j indexes a gene,  \(\bm{h}\) is substituted for the loadings or weights (for ASCA+ and PLS respectively) and VE is the variance explained, where $VE_{k}$ is the amount of variance explained by component k and $VE_{total}$ is the variance explained by the whole model. In ASCA+ we calculate the VIP scores in the latent space of \(\bm{Y}_{\alpha + \alpha\beta}\) as described in @thielASCAAPCAExtensions2017. In PLS we use the VE from the latent space of \(\bm{B}\) using the weights ($\mathbf{U}$) from the latent space of \(\bm{Y}\).

@@

```{r, echo = FALSE}
#### VST inverse #######
#mean/sd relationship plots
##Power transformation of counts/concentration matrix
m_sd_n <- M_SD(X, meta)

plot_normal <- m_sd_n[[3]]
```

```{r, echo=FALSE, include=FALSE}
db <- cbind.data.frame("Sample" = c(1:24),meta[,1:2],X)
db[,1:3] <- apply(db[,1:3],2,as.character)
mc <- reshape2::melt(db)
colnames(mc)[4:5] <- c("Feature","Expression")
```

```{r, echo = FALSE}
dist_norm <- ggplot(mc, aes(x = Expression)) + geom_density(alpha = 0.4)+
  scale_y_continuous(labels = percent_format()) +
  theme_bw()+
  ggtitle("Core Distribution")
```

```{r, echo=FALSE}
dist_nb <- ggplot(mc, aes(x = Expression)) + geom_density(alpha = 0.4)+
  scale_y_continuous(labels = percent_format()) +
  scale_x_continuous(guide = guide_axis(check.overlap = TRUE)) +
  theme_bw()+
  ggtitle("NB transformed")

# (dist_norm + plot_normal)  # no noise data so cannot plot mean sd relationships here
```

@@

```{r, echo = FALSE}
# comb_terms <- ASCA_combine_terms(res_ASCAplus_nb, c("growth_condition","growth_condition:time"))
# svd_test <- ASCA_svd(comb_terms$decomposition$`growth_condition+growth_condition:time`)
#

@@

######

SL_plot <- function(PCD,meta){

  a_c.PCD <- PCD
  meta <- as.data.frame(meta)
  a_c.PCD$x <- cbind.data.frame(time = meta$time, growth_condition = meta$growth_condition, a_c.PCD$x)

>>>>

@@

# Introduction
Plant transcriptome studies typically involve the use of designed experiments which aim to induce variation in pathways of interest (POIs) by controlling one or more experimental factors. This can be done through, for example, varying the level of certain essential nutrients, or knocking out/down a gene within a POI. To explore which pathways are affected by the intervention, gene expression can be measured through RNA sequencing (RNAseq). Analysis of RNAseq data generally adheres to one of two approaches: differential expression analysis (DE; @andersDifferentialExpressionAnalysis2010; @robinsonEdgeRBioconductorPackage2010) or coexpression analysis (CoE; @langfelderWGCNAPackageWeighted2008; @tzfadiaCoExpNetVizComparativeCoExpression2016). DE is used to determine which genes are expressed differently between two or more experimental conditions, highlighting known or novel pathways affected by the experimental conditions. CoE, on the other hand, aims at discovering genes that are part of a partially characterized POI. Here, the expression profiles of known genes of the POI (termed "baits") are used to detect novel pathway members with similar expression patterns (here termed "spikes").

```{r nettype, fig.cap= c("Pathway types exemplified, POI; pathway of interest, RP; related pathway, URP; unrelated pathway. A represents hypothetical biological pathways, B highlights the problem with detecting these pathways when dealing with data from designed experiments.")}

a <- knitr::include_graphics("networktypes.png")
a
```

Figure \@ref(fig:nettype) outlines pathway nomenclature that will be used throughout this text. *A* represents a true hypothetical pathway including a pathway of interest (POI), related pathway (RP) and an unrelated pathway (URP). All three types of pathway are assumed to be affected (dis-)proportionally by the experimental design. Note; URPs are not necessarily completely unlinked to POIs or RPs, in the biological sense, but are assumed here to be several degrees away. Far enough that the effect can be approximated orthogonally to the coexpression, when modelling. When dealing with experimentally designed data, we have to consider two key sources of variance: between-group variance and within-group variance. Between-group variance pertains to the differences between experimental groups (or conditions). This type of variance is of interest when conducting DE studies. Within-group variance, on the other hand, is the variance within experimental groups, this is not interesting for DE studies but may contain information useful to CoE. It is the excessive between-group variance that leads to the densely connected nature of *B* from \@ref(fig:nettype). Figure \@ref(fig:withinbetween) exemplifies between-group and within-group driven correlations from a real dataset (@haiderTranscriptomeAnalysisPhosphate2023). The plots show the relationship between two genes of the POI (*A* and *C*) and the relationship between two genes that are not part of the same pathway (*B* and *D*). The total correlations in both *A* and *B* are high but correlation in *B* is driven by the between-group variation and not by correlations within experimental groups. The within-group correlation is lower than the total in both cases, but there is a clear trend in *C*: P- condition in red. The within group correlation (r) is higher for the first transcript pair (*C*) than the second (*D*), which makes it possible to discern between genes within a POI or RP and genes that both respond similarly to a treatment (URP). Hence, we need to distinguish *total CoE* which is calculated on the total variance and *within-group CoE* where the focus is on the shared variance within experimental groups.

@@

@@

```{r, echo = FALSE, include=FALSE}
META <- as.data.frame(meta[,1:2])
META$growth_condition[which(META$growth_condition == "1")] <- "P+"
META$growth_condition[which(META$growth_condition == "-1")] <- "P-"

@@

What is notable here is the replicate specific patterns that are apparent amongst the SL genes and largely absent between the SLs (purple) and the non-SL differential genes (green). For example, at time point 3,  most SL genes are expressed lowest in replicate 1 and highest in replicate 3, while at time point 7 replicate 2 has a higher SL gene expression. These within-group patterns are either not shared at all between the SL genes and the other DE genes or to a much lesser extent. This is due to the other genes not directly being part of the same pathway. There are of course other pathways and thus there are still some subtle replicate patterns in the non-SL genes but this effect is much clearer within our pathway of interest. Figure \@ref(fig:ASCA-1) shows the output of the ASCA decomposition in which the PCA scores and loadings of the combined effect of the  Growth condition factor and the interaction of Growth condition and Time is explored. The score plot shows that the replicates of each condition are clustered while the difference between the conditions is much larger. The loading plot shows that although the 9 SL genes are mostly on the outside of the plot, many other genes are as well. Thus many differentially expressed genes are found important that are not part of the SL pathway, in other words ASCA does not detect SL (POI) genes from others that are similarly or more differentially expressed but not part of the SL pathway (URPs).

This marks the goal of this paper, which is to detect within-group coexpressed genes (POI/RP) with a set of baits preferentially to other differentially expressed genes that are not part of the same pathway (URPs, with no shared within-group structure).

@@

```{r Dataset1-Overview, echo = F, fig.cap=c("Real data overview. RNAseq of rice root. Heatmap of SL pathway and example highly differential genes. Data and preprocessing from @haiderTranscriptomeAnalysisPhosphate2023, genes autoscaled i.e. blue indicates no/low expression and red indicates higher expression.")}

@@

```{r ASCA-1, echo = F, out.width = '100%', fig.cap=c("Scores and loadings of the ASCA model of the combined effect (condition + condition:time interaction), model indicates samples and genes relative positions in the designed variance.")}
knitr::include_graphics("rice_s_l.png")

@@

The simulations are built upon the results of the ASCA analysis of the combined effect of growth condition and the interaction of growth condition and time (see Figure \@ref(fig:ASCA-1) for overview). The simulated data is created by simulating score and loading values (as in Figure \@ref(fig:simOverview)) for the combined effect ($\alpha + \alpha\beta$) ) explaining the difference in time profiles of the P+ and P- condition and the main time effect $\beta$, as well as, structured ambient variance ($S$). This is to mimic variance due to gene expression as a function of growth condition and time as well as variance due to coexpression of genes in the same  pathway on top of this experimentally induced DE variance.

@@

The first simulation aims to determine how well the methods perform to find coexpressed genes in the presence of a dominant between-group variance structure and increasing numbers of differential URP genes (genes with high between-group correlation but no within-group correlation). The expression levels of differential transcripts is modeled with the scores and loadings of the growth condition and its interaction with time; termed here the combined effect ( $\mathbf{T}_{\alpha + \alpha\beta}$ and $\mathbf{P}_{\alpha + \alpha\beta}^T$ ). The time effect is modeled with scores and loadings ( $\mathbf{T}_{\beta}$  and $\mathbf{P}_{\beta}^T$ ). Also included is some random measurement noise, this is modeled as a matrix $\mathbf{E}$ with normally distributed random values $e_{ij} = \mathcal{N}(0,\sigma^2)$.

On top of the between-group effect there is an additional correlation structure applied to the baits and spikes to create a model of coexpression that is independent to the variance induced by the experimental factors ($\mathbf{X}_S = \mathbf{T}_S\mathbf{P}_S^T$). This $\mathbf{X}_S$ matrix contains two component sources of variation which links the baits and spikes. In this simulation we have a set of 4 baits that have positive loadings for the combined effect along with a variable number of other genes - "Combined Effect Responders", which are the genes from a simulated URP (indicated by the yellow bar in Figure \@ref(fig:simOverview)). For the structured ambient variance; 2 random vectors in \(\bm{T}_S\) (each column representing an effect caused by ambient conditions) which our baits and spikes contribute to (i.e. green and black bars in Figure \@ref(fig:simOverview) are non zero in \(\bm{P}_S\)). Similarly to the real data, the gene expression values for each experimental condition are simulated with three independent replicates per time-condition-combination. Each simulated data $\mathbf{Y}$ set thus consists of 24 samples and 2000 genes.

@@

The scores matrices $\mathbf{T}_{\alpha + \alpha\beta}$ and $\mathbf{T}_{\beta}$ each consist of two orthogonal score vectors and their relative levels are indicated by the colour. The different colours in Figure \@ref(fig:simOverview) represent that the scores are orthogonal to each other. Each small block consists of the same score value for each of the replicates within that condition. Loadings $\mathbf{P}_{\alpha + \alpha\beta}^T$ and $\mathbf{P}_{\beta}^T$  are simulated to create the baits and spikes. As displayed in Figure \@ref(fig:simOverview), the green bar represents 4 baits (or known pathway genes), the black bar represents the 12 spikes (“unknown” genes from POI or RP), yellow represents differential URP genes, and purple; a set of differential URPs with an alternative treatment response.

In $\mathbf{P}_{\beta1}^T$ the baits, spikes and other DE genes (all genes from POI, RP and URP) along with the remaining first half have similar positive values (represented by light grey) where the other 1000 have positive values in $\mathbf{P}_{\beta2}^T$ (grey), these are created to emulate the reality of the real data in which many genes are slightly correlated with eachother through the course of development, as well as to account for some nonlinearities which are otherwise ignored by our model. The remaining elements of $\mathbf{P}_{\alpha + \alpha\beta}^T$ and $\mathbf{P}_{\beta}^T$ are populated with zeroes. The $\mathbf{T}$ matrices represent the simulated scores for their respective factors. For blues, reds, greys and golds we have 4 different shades representing 4 unique values per factor level (total 8 unique values where reds and golds are sign flipped blues and greys respectively). Similarly oranges and pinks represent a growth condition independent time profile with four unique values. $\mathbf{E}$ represents random unstructured technical noise and the additional $\bm{1\mu}^T$ term in \@ref(eq:Sim2-bp) represents a mean value for each gene.

```{r simOverview, echo=FALSE, fig.cap="Coexpression simulation, simplified overview. Colours correspond to the same genes in the different loading vectors (*P*). Black; baits (known POI genes), green; spikes (unknown POI genes), yellow; DE URP genes, purple DE URP genes with a different profile. In the score vectors (T); Blues, reds, greys and golds represent experimental conditions, oranges and pinks; time. Each of the textures in the ambient/residual part represent a distinct normally distributed set of random values.", out.width = '100%'}
knitr::include_graphics("Sim_SL_diagram_simplified.png")
```

\begin{equation}
\begin{aligned}
\bm{Y} &= \bm{1\mu}^T + \bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T + \bm{T}_{\beta}\bm{P}_{\beta}^T + \bm{T}_{S}\bm{P}_{S}^T + \bm{E} \\
\text{where for each element}\\
&\bm{T}_{S_{ik}} \sim \mathcal{N}(0,1), \quad \bm{P}_{S_{jk}}^{\text{baits,spikes}} \sim \mathcal{N}(0.5,0.2) \\
\text{and:} &\\
\bm{P}_{S}^{\text{rest}} &= 0
\end{aligned}(\#eq:Sim2-bp)
\end{equation}

The issue to address is the presence of many differential URPs; those that have strong between group correlations but no within group correlation. Here these differential genes relate  to the yellow bar in Figure \@ref(fig:simOverview). We incrementally increase the number of non-zero values and decrease the number of 0 values, which in figure \@ref(fig:simOverview) relates to increasing the size of the yellow bar, as well as varying the size (as quantified by the Frobenius norm) of the combined effect matrix through the scalar parameter *d* below;

@@

@@

In order to evaluate the methods and indicate the levels of variance required for optimal performance we conduct a larger scale assay. Here we fix the number of DE URP genes to 30 which corresponds to the yellow bar size and the size of the within correlation effect matrix (\(\|\bm{T}_S\bm{P}_S^T\|_F = 0.025\)). We investigate a range of values for sizes of between variance (*d*) and random noise (*l*) across random structures (which correspond to repeated simulations of randomised values for all $\mathbf{T}$ matrices). Parameters were adjusted according to \@ref(eq:Test-2) on 20 randomised $\mathbf{T}$ matrices.

@@

@@

@@

Pearson's correlation coefficient is calculated between all columns of \(\bm{Y}\). For ranked correlations Pearson's *r* is calculated between each bait and all other genes. Ranks are assigned to all non-baits transcripts based on *r* with the lowest rank for the highest correlation. These ranks are averaged across all baits, similarly to @poretskyMutRankShinyWebapplication2020.

<!-- For example the correlation between a SL gene $j_{sl}$ and a non-SL gene $j_q$ is calculated as: -->

<!-- \begin{equation} -->
<!-- \begin{aligned} -->
<!--   r_{j_{sl}j_{q}} = \frac{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl}) (y_{ij_{q}} - \bar{y}_j_{q})}{\sqrt{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl})^2 \sum_{i=1}^{I} (y_{ij_{q}} - \bar{y}_j_{q})^2}} -->
<!-- \end{aligned}(\#eq:Pearson) -->
<!-- \end{equation} -->

@@

<!-- In this work we calculate the model: -->

### PLS regression

Partial Least Squares (PLS2) is a multivariate regression technique used for modeling relationships between two sets of variables. Here we split our data matrix $\mathbf{Y}$ into sub-matrices; $\mathbf{Y} = \left[\mathbf{\tilde{Y}} : \mathbf{B}\right]$

The response matrix \( \mathbf{B} \) contains only the baits (relating to the green bar in the loadings of Figure \@ref(fig:simOverview)) and the predictor matrix \( \mathbf{\tilde{Y}} \) contains the rest of the data, this can then be expressed as:

\begin{equation}
\begin{aligned}
\mathbf{\tilde{Y}} = \mathbf{T}\mathbf{P}^T + \mathbf{E}_\tilde{Y} \\
\text{and} \\
\mathbf{B} = \mathbf{T}\mathbf{Q}^T + \mathbf{E}_B \\
\end{aligned}
\end{equation}

This model is estimated in our work using the SIMPLS algorithm (@dejongSIMPLSAlternativeApproach1993) as follows:

1. Initialize \( \mathbf{\tilde{Y}}_0 \) to \( \mathbf{\tilde{Y}} \) and \( \mathbf{B}_0 \) to \( \mathbf{B} \).
2. Iteratively perform the following steps for \( k = 1, 2, \ldots \):
   a. Compute the singular value decomposition (SVD) of \( \mathbf{\tilde{Y}}_{k}^{T} \mathbf{B}_{k} \) to obtain the left and right singular vectors \( \mathbf{u}_k \) and \( \mathbf{v}_k \) respectively.
   b. Project \( \mathbf{\tilde{Y}}_{k} \) and \( \mathbf{B}_{k} \) onto the first left and right singular vectors (\( \mathbf{u}_k \), \( \mathbf{v}_k \)) to obtain the scores \( \bm{t}_k = \mathbf{\tilde{Y}}_{k} \mathbf{u}_k \) and \( \bm{w}_k = \mathbf{B}_{k} \mathbf{v}_k \).
   c. Perform linear regression of \( \mathbf{\tilde{Y}}_k \) and \( \mathbf{B}_k \) onto their corresponding scores \( \bm{t}_k \) and \( \bm{w}_k \) to obtain the respective loadings \( \bm{p}_k \), and \( \bm{q}_k \).
   d. Deflate \( \mathbf{\tilde{Y}}_k \) and \( \mathbf{B}_k \) by removing the variance explained by the $k^{th}$ set of scores in the space of $\mathbf{\tilde{Y}}$\( \bm{t}_k \) and use the deflated matrices as inputs for the next iteration.

\begin{equation}
\begin{aligned}
\mathbf{\tilde{Y}_{k}} = \mathbf{\tilde{Y}}_{k-1} - \mathbf{t}_{k}\mathbf{p}^T_{k} \\
\text{and} \\
\mathbf{B_{k}} = \mathbf{B}_{k-1} - \mathbf{t}_{k}\mathbf{q}^T_{k} \\
\end{aligned}
\end{equation}

PLS2 aims to maximize the covariance between the scores \( \bm{t}_k \) and \( \bm{w}_k \) at each iteration. In the following simulations and application the number of components calculated by PLS2 is set to 2 to create an interpretable model.

@@

$\text{VIP}_{j} = \sqrt{\frac{ \sum_{k=1}^{K} \mathbf{H}^2_{jk} \cdot \text{VE}_{k}}{\text{VE}_{total} \cdot K}}$

Where k=1...K represents the number of components in the respective models, j indexes a gene,  \(\bm{H}\) is substituted for the loadings or weights (for ASCA+ and PLS respectively) and VE is the variance explained, where $VE_{k}$ is the amount of variance explained by component k and $VE_{total}$ is the variance explained by the whole model. In ASCA+ we calculate the VIP scores in the latent space of \(\bm{Y}_{\alpha + \alpha\beta}\) as described in @thielASCAAPCAExtensions2017. In PLS we use the VE from the latent space of \(\bm{B}\) using the weights ($\mathbf{U}$) from the latent space of \(\bm{Y}\).

@@

@@

@@
